\documentclass[12pt, leqno, british]{amsart}
\input{../preamble}
\addbibresource{../bibliography.bib}
\externaldocument[M-]{../Lecture-notes}

\author{Nicolas Daans}
\address{Charles University, Faculty of Mathematics and Physics, Department of Algebra, Sokolov\-sk\' a 83, 18600 Praha~8, Czech Republic.}
\email{nicolas.daans@matfyz.cuni.cz}

\begin{document}

\section{Lecture 1}
\subsection{Bilinear and quadratic forms}
Let always $K$ be a field, $n \in \nat$.
\begin{defi}
A \emph{symmetric bilinear space over $K$}\index{symmetric bilinear!space} is a pair $(V, B)$ where
\begin{itemize}
\item $V$ is a finite-dimensional vector space over $K$, and
\item $B : V \times V \to K$ is a symmetric and bilinear map, i.e.~for all $x, x', y \in V$ and $a \in K$ we have
\begin{align*}
B(x, y) &= B(y, x), \\
B(x+x',y) &= B(x, y) + B(x', y), \\
B(ax, y) &= aB(x, y).
\end{align*}
\end{itemize}
We call the map $B$ a \emph{symmetric bilinear form on $V$}.\index{symmetric bilinear!form}
We define the dimension of $(V, B)$ to be the dimension of $V$, and denote this by $\dim(V, B)$ or simply $\dim B$.

Let $n = \dim (V, B)$. Given a basis $\mc{B} = (e_1, \ldots, e_n)$ of $V$, we define $\mc{M}_\mc{B}(B) = [B(e_j, e_i)]_{i, j=1}^n$, which we call \emph{the matrix of $(V, B)$ with respect to $\mc{B}$}.
\end{defi}
\begin{prop}
Let $V = K^n$ and let $\mf{B} = (e_1, \ldots, e_n)$ be the canonical basis.
Let $B$ be a symmetric bilinear form on $V$.
For column vectors $x = [x_1 \ldots x_n]^T$ and $y = [y_1, \ldots, y_n]^T$ we have
$$ B(x, y) = x^T\mc{M}_{\mc{B}}(B)y.$$
\end{prop}
\begin{proof}
This is clear from the bilinearity of $B$.
\end{proof}
\begin{defi}\label{D:QF}
A \emph{quadratic space over $K$}\index{quadratic!space} is a pair $(V, q)$ where
\begin{itemize}
\item $V$ is a finite-dimensional vector space over $K$, and
\item $q : V \to K$ is a map satisfying the following:
\begin{enumerate}
\item $\forall a \in K, \forall x \in V : q(ax) = a^2q(x)$,
\item the map $$\mf{b}_q : V \times V \to K : (x, y) \mapsto q(x+y) - q(x) - q(y) $$
is a symmetric bilinear form on $V$.
\end{enumerate}
\end{itemize}
We call the map $q$ a \emph{quadratic form on $V$}\index{quadratic!form}, and $\mf{b}_q$ its \emph{polar form}\index{polar form}.
We define the dimension of $(V, q)$ to be the dimension of $V$, and denote this by $\dim(V, q)$ or simply $\dim q$.
\end{defi}
\begin{defi}
Let $(V, B)$ and $(V', B')$ be symmetric bilinear spaces over $K$.
An isomorphism of $K$-vector spaces $I : V \to V'$ is called an \emph{isometry between $(V, B)$ and $(V', B')$}\index{isometry} if, for all $v, w \in V$, one has $B(v, w) = B'(I(v), I(w))$.
Similarly, given quadratic spaces $(V, q)$ and $(V', q')$ over $K$, an isomorphism of $K$-vector spaces $I : V \to V'$ is called an \emph{isometry between $(V, q)$ and $(V', q')$} if, for all $v \in V$, one has $q(v) = q'(I(v))$.

We call two symmetric bilinear spaces $(V, B)$ and $(V, B')$ (respectively two quadratic spaces $(V, q)$ and $(V', q')$) \emph{isometric}, which we denote by $(V, B) \cong (V', B')$ (respectively $(V, q) \cong (V', q')$) if there exists an isometry between them.
\end{defi}

Traditionally, a quadratic form over $K$ is often defined to be a homogeneous polynomial of degree $2$ over $K$.
\Cref{D:QF} can be seen as a coordinate-free version of this, as the following proposition indicates.
\begin{prop}\label{P:QF-coordinates}
Let $n \in \nat$ and let $f \in K[X_1, \ldots, X_n]$ be a homogeneous polynomial of degree $2$. The map
$$ q_f : K^n \to K : (x_1, \ldots, x_n) \mapsto f(x_1, \ldots, x_n)$$
is a quadratic form on $K^n$.

Conversely, given a quadratic space $(V, q)$ of dimension $n$, there exists a homogeneous degree $2$ polynomial $f \in K[X_1, \ldots, X_n]$ such that $(V, q)$ is isometric to $(K^n, q_f)$.
\end{prop}
\begin{proof}
For the first part of the statement, one verifies that the defined map satisfies the conditions stated in \Cref{D:QF}.

The second part of the statement is left as an exercise.
\end{proof}
\begin{prop}\label{P:isometric-coordinate}
Let $f, g \in K[X_1, \ldots, X_n]$ be homogeneous polynomials of degree $2$.
The quadratic spaces $(K^n, q_f)$ and $(K^n, q_g)$ are isometric if and only if there exists $C \in \GL_n(K)$ such that
$$ f\left(\left[ x_1 \ldots x_n \right]^T\right) = g\left((\left[ x_1 \ldots x_n\right] C)^T\right)$$
for all $x_1, \ldots, x_n \in K$.
\end{prop}
\begin{proof}
Exercise.
\end{proof}
\begin{eg}\label{E:hyperbolic-plane-isometry}
Suppose $\charac(K) \neq 2$. Let $f(X_1, X_2) = X_1 \cdot X_2$ and $g(X_1, X_2) = X_1^2 - X_2^2$.
We observe that
$$ g\left(\frac{X_1 + X_2}{2}, \frac{X_1 - X_2}{2}\right) = f(X_1, X_2) $$
and thus, in view of \Cref{P:isometric-coordinate}, that $(K^2, q_f) \cong (K^2, q_g)$, with
\begin{displaymath}
C = \begin{bmatrix}
\frac{1}{2} & \frac{1}{2} \\ \frac{1}{2} & \frac{-1}{2}
\end{bmatrix}.
\end{displaymath}
\end{eg}
We saw that, to a quadratic form $q$, one can associate a symmetric bilinear form $\mf{b}_q$ on the same space.
It is also possible to obtain a quadratic form from a symmetric bilinear form: if $(V, B)$ is a symmetric bilinear space, then
$$ q_B : V \to K : v \mapsto B(v, v)$$
is easily seen to be a quadratic form.
If $\charac(K) \neq 2$, then these two operations are each others inverses (up to scaling by $\frac{1}{2}$), and hence the studies of quadratic and symmetric bilinear forms over $K$ are essentially the same:
\begin{prop}\label{P:quadratic-vs-SBS}
Assume $\charac(K) \neq 2$. Let $(V, q)$ be a quadratic space.
Then $q$ is equal to the quadratic form associated to the form $\frac{1}{2}\mf{b}_q$.
Conversely, if $(V, B)$ is a symmetric bilinear space, then $B$ is equal to $\frac{1}{2}\mf{b}_q$ where $q = q_B$.
\end{prop}
\begin{proof}
This is a straightforward computation.
\end{proof}
Over fields of characteristic $2$, one can still associate to each quadratic form a symmetric bilinear form and to each symmetric bilinear form a quadratic form as before, but these operations are not invertible.
In fact, one needs to make an entirely separate study of quadratic and symmetric bilinear forms! We refer the interested reader to \autocite[Chapters I and II]{ElmanKarpenkoMerkurjev}.

We now go on to study basic properties of quadratic forms.
\begin{defi}\label{D:isotropic-represents-universal}
Let $(V, q)$ be a quadratic space over $K$.
\begin{itemize}
\item We call $q$ \emph{isotropic}\index{isotropic} if there exists $v \in V \setminus \lbrace 0 \rbrace$ such that $q(v) = 0$, or \emph{anisotropic}\index{anisotropic|see{isotropic}} otherwise.
\item Given $a \in K^\times$, we say that \emph{$q$ represents $a$}\index{representation!of an element by a form} if $\exists v \in V$ with $a = q(v)$.
We write
$$ D_K(q) = \lbrace a \in K^\times \mid \exists v \in V : a = q(v) \rbrace.$$
If $D_K(q) = K^\times$, we say that $q$ is \emph{universal}\index{universal}.
\end{itemize}
\end{defi}
\begin{egs}\label{E:hyp} \
\begin{enumerate}
\item Let $f(X_1, X_2) = X_1 \cdot X_2$. Then $q_f$ is isotropic, since $f(1, 0) = 0$.
$q_f$ is also universal, since, $f(1, a) = a$ for any $a \in K^\times$.
\item Let $f(X_1, X_2) = X_1^2 + X_2^2$.
$q_f$ is isotropic if and only if $-1$ is a square in $K$. $D_K(q_f)$ is the set of elements of $K$ which are a sum of two squares.
\item Let $f(X_1, X_2) = (X_1 + X_2)^2$. Then $q_f$ is isotropic since $f(1, -1) = 0$.
$D_K(q_f)$ consists of those elements of $K$ which are squares.
\end{enumerate}
\end{egs}
The last example is somewhat peculiar: the quadratic form $q_f$ with $f(X_1, X_2) = (X_1 + X_2)^2$ is of dimension $2$, but after a base change, one of the variables disappears. Indeed,
$$ f\left(X_1 - X_2, X_2\right) = X_1^2.$$
We will often want to exclude from our study quadratic forms which have this property.

For a $K$-vector space $V$, we denote by $V^\ast$ the dual space, i.e.~the space of linear maps $V \to K$.
Recall that $\dim(V^\ast) = \dim(V)$.
\begin{prop}\label{P:nonsingular-characterisations}
Let $(V, B)$ be a symmetric bilinear space.
Let $\mc{B}$ be a basis for $V$.
The following are equivalent.
\begin{enumerate}[(a)]
\item $\forall v \in V \setminus \lbrace 0 \rbrace$, $\exists w \in V$ : $B(v, w) \neq 0$,
\item The map $V \to V^\ast : v \mapsto (w \mapsto B(v, w))$ is a $K$-isomorphism.
\item The matrix $M_\mc{B}(B)$ is invertible.
\end{enumerate}
\end{prop}
\begin{proof}
Exercise.
\end{proof}
\begin{defi}
We call a symmetric bilinear space $(V, B)$ \emph{nonsingular}\index{nonsingular} if the above equivalent conditions hold.
We call a quadratic space $(V, q)$ nonsingular if its polar form is nonsingular.
Otherwise, we call the space \emph{singular}.
We use the same terminology for the symmetric bilinear and quadratic forms themselves.
\end{defi}
We now show that, at least over fields of characteristic not $2$, singular forms are precisely those for which, after a base change, one of the variables disappears.
\begin{prop}\label{P:nonsingular-polynomials}
Let $(V, q)$ be a quadratic space over $K$ and $v \in V$. Consider the statements
\begin{enumerate}[(a)]
\item\label{it:singular-1} $\mf{b}_q(v, w) = 0$ for all $w \in V$,
\item\label{it:singular-2} for all $w \in V$ we have $q(w + v) = q(w)$.
\end{enumerate}
We have that \eqref{it:singular-2} $\Rightarrow$ \eqref{it:singular-1} in general.
If $\charac(K) \neq 2$, then \eqref{it:singular-1} and \eqref{it:singular-2} are equivalent.

In particular, it follows that, if $\charac(K) \neq 2$, a quadratic space $(V, q)$ is singular if and only if there exists $v \in V \setminus \lbrace 0 \rbrace$ such that for all $w \in V$ we have $q(w+v) = q(w)$.
\end{prop}
\begin{proof}
If \eqref{it:singular-2} holds, then $q(v) = q(v + 0) = q(0) = 0$, whence for any $w \in V$ we have $\mf{b}_q(v, w) = q(v + w) - q(v) - q(w) = 0$. %Thus, $(V, q)$ is singular.

Now assume that $\charac(K) \neq 2$ and \eqref{it:singular-1} holds.
%Then there exists $v \in V$ such that $\mf{b}_q(v, w) = 0$ for all $w \in V$.
Then in particular $0 = \mf{b}_q(v, v) = 2q(v)$ and thus $q(v) = 0$.
It follows that, for any $w \in V$, we have $q(v + w) = q(v) + q(w) + \mf{b}_q(v, w) = q(v)$, so \eqref{it:singular-2} holds.
\end{proof}
If $\charac(K) \neq 2$, a nonsingular quadratic form over $K$ is also called \emph{regular} or \emph{nondegenerate}.
Note that, if $\charac(K) = 2$, these terms have more specialised, distinct meanings.

\begin{rem}
So far, I have been somewhat careful in making the distinction between a symmetric bilinear/quadratic \textit{space} and a symmetric bilinear/quadratic \textit{form}.
This makes notation and speaking somewhat heavy. I will in the future often simply refer to the forms themselves, taking the convention that a symmetric bilinear/quadratic space `knows' its domain.
\end{rem}

\subsection{Orthogonality and diagonalisation}
\begin{defi}
Let $(V, B)$ be a symmetric bilinear space.
Let $v, w \in V$.
We say that $v$ and $w$ are \emph{orthogonal (with respect to $B$)}\index{orthogonal} if $B(v, w) = 0$.
We write $v \perp w$.

Let $v \in V$ and $M \subseteq V$.
We say that $v$ is \emph{orthogonal to $M$ (with respect to $B$)} if $B(v, w) = 0$ for all $w \in M$.
We write $v \perp M$.
Similarly, given $M' \subseteq V$, we say that $M$ is \emph{orthogonal to $M'$ (with respect to $B$)} if $B(v, w) = 0$ for all $v \in M$ and $w \in M'$, and write $M \perp M'$.

We write
$$ M^\perp = \lbrace v \in V \mid \forall w \in M : B(v, w) = 0 \rbrace$$
and call it the \emph{orthogonal space of $M$} - note that it is always a subspace of $V$. We write $v^\perp$ instead of $\lbrace v \rbrace^\perp$.

If $U \subseteq V$ is a subspace and $V = U \oplus U^\perp$, we call $U^\perp$ an \emph{orthogonal complement of $U$ in $V$}.
\end{defi}
Observe that a symmetric bilinear space $(V, B)$ is by definition nonsingular if and only if $V^\perp = \lbrace 0 \rbrace$.

\begin{prop}\label{P:dim-duality}
Let $(V, B)$ be nonsingular, $U \subseteq V$ a subspace. Then
$$ \dim U + \dim U^\perp = \dim V \quad\text{and}\quad (U^\perp)^\perp = U.$$
\end{prop}
\begin{proof}
Consider the $K$-linear maps
\begin{align*}
\varphi_1 &: U^\perp \to V^\ast : v \mapsto (w \mapsto B(v, w)) \\
\varphi_2 &: V^\ast \to U^\ast : f \mapsto f\vert_U.
\end{align*}
We observe that $\varphi_1$ is injective by the nonsingularity of $(V, B)$, that $\varphi_2$ is surjective, and that the image of $\varphi_1$ is precisely the kernel of $\varphi_2$ by definition of $U^\perp$.
As such, we compute that
\begin{align*}
\dim V &= \dim V^\ast = \dim(\Ker \varphi_2) + \dim(\Img \varphi_2) \\
&= \dim(\Img \varphi_1) + \dim U^\ast = \dim (U^\perp) + \dim(U)
\end{align*}
as desired.

For the second statement, observe that we trivially have $U \subseteq (U^\perp)^\perp$, but that, by the first claim, $\dim(U) = \dim((U^\perp)^\perp)$, whence $U = (U^\perp)^\perp$ as desired.
\end{proof}
We now define an operation on the set of quadratic spaces over $K$.
\begin{prop}\label{P:orth-sum}
Let $(V_1, q_1)$ and $(V_2, q_2)$ be quadratic spaces over $K$. Let $V = V_1 \times V_2$ and consider the map
$$ q : V \to K : (x, y) \mapsto q_1(x) + q_2(y). $$
Furthermore, consider the natural embeddings $\iota_1 : V_1 \to V : x \mapsto (x, 0)$ and $\iota_2 : V_2 \to V : x \mapsto (0, x)$.
We have that $(V, q)$ is a quadratic space, and $q$ is nonsingular if and only if both $q_1$ and $q_2$ are.
Furthermore, $i_1(V_1) \perp i_2(V_2)$ with respect to $\mf{b}_q$.
\end{prop}
\begin{proof}
Easy verification.
\end{proof}
\begin{defi}
Let $(V_1, q_1)$ and $(V_2, q_2)$ be quadratic spaces over $K$. We call the space $(V, q)$ defined in \Cref{P:orth-sum} the \emph{orthogonal sum of $(V_1, q_1)$ and $(V_2, q_2)$} and we denote the form $q$ by $q_1 \perp q_2$.
\end{defi}
\begin{prop}
Let $(V_i, q_i)$ and $(V_i', q_i')$ be quadratic spaces for $i = 1, 2, 3$. We have the following computation rules:
\begin{itemize}
\item $\dim(q_1 \perp q_2) = \dim(q_1) + \dim(q_2)$.
\item $q_1 \perp q_2 \cong q_2 \perp q_1$, and $q_1 \perp (q_2 \perp q_3) \cong (q_1 \perp q_2) \perp q_3$.
\item If $q_1 \cong q_1'$ and $q_2 \cong q_2'$, then $q_1 \perp q_1' \cong q_2 \perp q_2'$.
\end{itemize}
\end{prop}
\begin{proof}
Easy verifications.
\end{proof}
\begin{prop}\label{P:intrinsic-orth-sum}
Let $(V, q)$, $(V_1, q_1)$ and $(V_2, q_2)$ be quadratic spaces over $K$.
Then $q \cong q_1 \perp q_2$ if and only if there are $K$-subspaces $W_1$ and $W_2$ of $V$ with $W_1 \perp W_2$ with respect to $\mf{b}_q$, $V = W_1 \oplus W_2$, and such that $(W_i, q\vert_{W_i}) \cong (V_i, q_i)$ for $i = 1, 2$.
\end{prop}
\begin{proof}
Suppose that $\iota$ is an isomorphism $q_1 \perp q_2 \to q$ and let $W_1$ and $W_2$ be the images under this isomorphism of $V_1 \times \lbrace 0 \rbrace$ and $\lbrace 0 \rbrace \times V_2$ respectively. One verifies easily that these are as desired.

Conversely, assume that $W_1$ and $W_2$ are subspaces of $V$ with $W_1 \perp W_2$, $V = W_1 \oplus W_2$, and such that $(W_i, q\vert_{W_i}) \cong (V_i, q_i)$ for $i = 1, 2$.
Without loss of generality, we may assume that $V_i = W_i$ and $q_i = q\vert_{W_i}$.
Let $\iota$ be the unique $K$-linear map $V \to V_1 \times V_2$ which maps a vector $w \in W_1$ to $(w, 0)$ and a vector $w \in W_2$ to $(0, w)$.
Clearly this is an isomorphism of $K$-vector spaces.
Consider an arbitrary vector in $V$, which we may write as $w_1 + w_2$ for $w_1 \in W_1$ and $w_2 \in W_2$.
Since $W_1 \perp W_2$, we have that $\mf{b}_q(w_1, w_2) = 0$. We compute that
\begin{align*}
q(w_1 + w_2) &= q(w_1) + q(w_2) + \mf{b}_q(w_1, w_2) = q(w_1) + q(w_2) \\
&= q_1(w_1) + q_2(w_2) = (q_1 \perp q_2)(w_1, w_2) = (q_1 \perp q_2)(\iota(w_1 + w_2)).
\end{align*}
Hence $\iota$ is the desired isometry.
\end{proof}

We now discuss a special class of quadratic forms called diagonal forms. As it will turn out, in characteristic different from $2$, every quadratic form is isometric to a diagonal form (see \Cref{C:diagonalisation}).
\begin{defi}
Let $a_1, \ldots, a_n \in K$. We denote by $\langle a_1, \ldots, a_n \rangle_K$ the quadratic form
$$ K^n \to K : (x_1, \ldots, x_n) \mapsto \sum_{i=1}^n a_ix_i^2.$$
We call such a form a \emph{diagonal form}\index{diagonal form}.
If the field $K$ is clear from the context we might simply write $\langle a_1, \ldots, a_n \rangle$ instead of $\langle a_1, \ldots, a_n \rangle_K$.
\end{defi}
Note that $\langle a_1, \ldots, a_n \rangle_K \cong \langle a_1 \rangle_K \perp \ldots \perp \langle a_n \rangle_K$.
\begin{prop}\label{P:diagforms-singular}
Let $n \in \nat$ and $a_1, \ldots, a_n \in K$, let $q = \langle a_1, \ldots, a_n \rangle$.
If $\charac(K) \neq 2$, then $q$ is singular if and only if $a_i = 0$ for some $i \in \lbrace 1, \ldots, n \rangle$.
If $\charac(K) = 2$, then $q$ is singular as soon as $n \geq 2$.
\end{prop}
\begin{proof}
Exercise.
\end{proof}
\begin{prop}\label{P:diagonalisation}
Assume $\charac(K) \neq 2$.
Let $(V, q)$ be a quadratic space over $K$, $d \in K^\times$.
Then $d \in D_K(q)$ if and only if $q \cong \langle d \rangle \perp (V', q')$ for some quadratic space $(V', q')$.
\end{prop}
\begin{proof}
Clearly $d = d \cdot 1^2 + q'(0) \in D_K(\langle d \rangle \perp (V', q'))$ for any quadratic space $(V', q')$.

Conversely, assume that $d \in D_K(q)$.
Let $W$ be any subspace of $V$ such that $V = V^\perp \oplus W$.
Then $(W, q\vert_W)$ is nonsingular, and, in view of \Cref{P:nonsingular-polynomials}, we have $D_K(q\vert_W) = D_K(q)$.
We may thus restrict our quadratic form to $W$, and assume without loss of generality that $q$ is nonsingular.

Now take $v \in V$ with $q(v) = d$.
Set $U = v^\perp$.
We have $v \not\in v^\perp$ (since $\mf{b}_q(v, v) = 2d \neq 0$) and $\dim(U) = \dim(V) - 1$ by \Cref{P:dim-duality}, hence $V = Kv \oplus U$.
Clearly $q\vert_{Kv} \cong \langle d \rangle$, so $q \cong \langle d \rangle \perp (U, q\vert_U)$ in view of \Cref{P:intrinsic-orth-sum}.
\end{proof}
\begin{cor}\label{C:diagonalisation}
Assume $\charac(K) \neq 2$, let $(V, q)$ be a quadratic space over $K$ of dimension $n$.
Then there exist $a_1, \ldots, a_n \in K$ such that $q \cong \langle a_1, \ldots, a_n \rangle$.
\end{cor}
\begin{proof}
Apply \Cref{P:diagonalisation} inductively.
\end{proof}

\subsection{Exercises}
\begin{enumerate}
\item Complete the proofs of \Cref{P:QF-coordinates}, \Cref{P:isometric-coordinate}, \Cref{P:nonsingular-characterisations} and \Cref{P:diagforms-singular}.
\item Illustrate by an example that the implication \eqref{it:singular-1} $\Rightarrow$ \eqref{it:singular-2} in \Cref{P:nonsingular-polynomials} does not hold in general if $\charac(K) = 2$.
\item Consider the quadratic form on $\qq^3$ given by the following polynomial:
$$f(X_1, X_2, X_3) = 3X_1^2 + 6X_1X_2 + 3X_2^2 - X_2X_3.$$
Explicitly construct a diagonal quadratic form $q$ on $\qq^3$ such that $(\qq^3, q_f) \cong (\qq^3, q)$.
\end{enumerate}
\end{document}